A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-10 paper.tex
Whittle down discussion items

There are some interesting consequences to these definitions, exemplified
in Figure~\ref{fig-arg-in-pedigree}. Firstly,
there is
is no limitation on mating system, ploidy, or age structure. By making the
basic unit of analysis an individual's \emph{genomes} (i.e., two per individuals
for diploids), any pattern of inheritance among mixed ploidy can be
accommodated.
Secondly, because we only record the \emph{intervals} of inheritance
between genomes (and not the causal \emph{processes}),
arbitrary patterns of genetic inheritance such as gene conversions,
multiple recombinations and so on can be directly expressed.
Thirdly, ``local`` trees that we generate for a particular position on the
genome are not explicitly concerned with ``coalescence'' (nodes at which
one or more samples find a common ancestor); for example, node $e$
is unary in the left-most tree of Figure~\ref{fig-arg-in-pedigree},
denoting the fact that $e$ the ancestor of $a$ over the interval $(0, 2]$,
but no coalescence occurs in this interval (see section XXX for more
discussion). Fourth, recombination is modelled in terms of
its outcomes in terms of inheritance intervals among parental genomes;
thus, we will have a child node (the recombinant; e.g. node $a$
Figure~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
paternal genomes which this child has inherited segments from.
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the evolutionary processes of genetic inheritance and
recombination, and is agnostic to the details of those processes.

@@

ARGs are most often thought of as a \emph{retrospective} structure however,
defined with respect to some set of (usually, contemporary) sample genomes.
In this case we are not explicitly interested in the
entire interval of genome descendant $u$ inherited from ancestor $v$,
but rather the subset of that interval that is \emph{ancestral to our
sample}.
Fig~\ref{fig-ancestry-resolution}B shows the graph that we obtain when
reduce the annotation intervals to those that are ancestral to the
sample (note that node $q$ is now ommitted because the MRCA of the sample
has already been found for the full genome at node $p$).
Tracking such
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
is usually much more useful and, indeed, all that is possible to
have information about when inferring ARGs from contemporary samples.
We refer to a gARG with inheritance annotations of this type as
``sample resolved''.

@@

\section*{Computational efficiency and data interchange}
ARG inference methods have been successfully applied to vast datasets,
including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
from UK Biobank~\citep{bycroft2018genome}
and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
Even larger datasets are
available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
and it seems inevitable that ARGs will be inferred for them.
The efficiency with which we can store and analyse these inferred
ARGs is therefore of great importantance, and a key factor
in the ultimate success of downstream ARG-based methods.
Another key factor in how important ARG-based methods become
across the wide range of predicted application
% TODO more?
areas~\citep{rasmussen2014genome,harris2019database,
hejase2020summary,schaefer2021ancestral,
fan2022genealogical,wohns2022unified,harris2023using}
is how well inference methods and downstream analysis applications
interoperate. Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems of converting data from one programs
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
While current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package
% FIXME-  more explicit?
(e.g.\ Relate, ARGneedle)
this approach scales poorly in terms of software development
effort, and is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse
software ecosystem.

@@

@@

\item It is unfortunate that we are adding yet another meaning to the
term ARG here, but all we're really doing is making the implicit
assumptions that others have made (i.e., describing the output of
Relate, tsinfer etc as ARGs) and making them concrete. We need to have a
term to describe all of these things, and since people are already using
ARG for this, we might as well make it rigorous. We can then move away
from the unfruitful discussion about which methods infer a ``true'' or
``real'' ARG (none do, really, under the original definitions) and
instead start discussing the \emph{properties} of these ARGs. (See
next point also.)
\item Terminology based around an inferred ancestry being a ``true''
ARG is impoverished and misleading. The only reasonable interpretation
of a ARG being true or not is whether it's a sample from the coalescent.
Most inference methods (and all that scale well) do not. In any case,
the coalescent is just one model, and a highly idealised one. It is
exceedingly unlikely that the actual ARG describing the ancestry
of any biological population is ``true'' ARG in this sense. The terminology is
impoverished because it reduces the possibilities available
to something either being an ARG or not, whereas there are many different
interesting properties of these structures that we should be able to
discuss and compare.
\item tskit can represent any ARG.
\item Interchange is a real problem, and if we're going to realise the
potential for ARGs in genomics we must solve it. Efforts to standardise
phylogenetic networks haven't done very well (~\citep{cardona2008extended}
has 89 citations). The only real effort to standardise an ARG format
based on extending GraphML~\citep{mcgill2013graphml} has 11 citations.
GraphML~\citep{mcgill2013graphml} has some of the same key ideas
as tskit though, in particular allowing the annotation of ``live sites''
to an edge as a set of disjoint intervals that hold ancestral material.

\item There are real computational disadvantages to the Griffiths ARG
data structure. Having two types of node (three really, as we need
``sample'' nodes which are also treated differently) complicates algorithms,
and makes it difficult to reason about the structure of the ARG as a
graph. For example, there's some weirdness about mutations over
recombination nodes discussed by Gusfield [Ana?]

@@

\item Throughout, we considered ARGs defined for a single chromosome comprising
of a fixed number of discrete sites. Our argument applies equally well to ARGs
defined for several non-homologous chromosomes \citep{fearnhead2003ancestral,
koskela2019robust} and to idealised, continuous chromosomes.

\item The gARG structure as described here is sufficiently expressive
to model any form of homologous genetic transmission between individuals,
and is far more flexible than the classical eARG approach.
Phylogenetic networks~\citep{huson2010phylogenetic} are a more
general concept, and specific classes capture dynamics that
cannot be modelled using an ARG. For example,
reassortment networks model the exchange of
(non-recombining) genomic segments between viral particles inside a co-infected host,
and hybridisation networks, model the genetic material from
different species combining to create a hybrid.
Similarly Genome Graphs~\cite{novak2017genome} model complicated things
within species. Future work should look to capture all of these things
in a single structure.

@@

% TODO not clear this belongs in the discussion, but we'll definitely
% want to make some of these points, even if they are being repeated
% from earlier section.
It is difficult to imagine a practical calculation performed on an
ARG that does not involve either the sequential generation of trees
along the genome or the resolved ancestry segments.
Knowing the resolved segments in an ancestry resolved gARG allows
us to perform many calculations without explicitly generating
the trees, and lets us work with the ancestry on a genome-by-genome
basis. We give three examples here. Firstly, generating mutations
on edges can be performed very quickly on an ancestry resolved
gARG, since we only need to examine $O(|E|)$ edges rather than
$O(nt)$ tree branches (assuming $n$ samples and $t$ marginal trees)
~\citep{baumdicker2021efficient}. If the ancestry were not resolved,
we would generate a large number of mutations on areas of the edge that
are non-ancestral. Secondly, computing the likelihood of a given
ARG realisation under the coalescent with recombination model
requires that we essentially recapitulate the state of Hudson's
algorithm (see XXX section) at each node in
the graph~\citep{baumdicker2021efficient}. Having resolved ancestry
substantially simplifies this process. Thirdly, computing
[FIXME describe what we did in tsdate]~\citep{wohns2022unified}.

Sequentially generating the marginal trees along the genome
is also fundamental, and is necessary whenever we need to
perform calculations that are contingent on more than just the
isolated properties of an edge. \cite{kelleher2016efficient}
showed how all trees can be sequentially generated in
constant time per tree transition in a fully simplified gARG.
Furthermore, we can easily reason about how tree topologies
change (and stay the same), leading to efficient algorithms
for computing population genetic
statistics~\citep{kelleher2016efficient,kelleher2018efficient},
implementing the Li and Stephens
model~\citep{kelleher2019inferring,wohns2022unified}
and likely many more.

>>>>

There are some interesting consequences to these definitions, exemplified
in Figure~\ref{fig-arg-in-pedigree}. Firstly,
there is
is no limitation on mating system, ploidy, or age structure. By making the
basic unit of analysis an individual's \emph{genomes} (i.e., two per individuals
for diploids), any pattern of inheritance among mixed ploidy can be
accommodated.
Secondly, because we only record the \emph{intervals} of inheritance
between genomes (and not the causal \emph{processes}),
arbitrary patterns of genetic inheritance such as
gene conversions,
multiple recombinations
across different chromosomes~\citep{fearnhead2003ancestral,koskela2019robust},
and so on can be directly expressed.

Thirdly, ``local`` trees that we generate for a particular position on the
genome are not explicitly concerned with ``coalescence'' (nodes at which
one or more samples find a common ancestor); for example, node $e$
is unary in the left-most tree of Figure~\ref{fig-arg-in-pedigree},
denoting the fact that $e$ the ancestor of $a$ over the interval $(0, 2]$,
but no coalescence occurs in this interval (see section XXX for more
discussion). Fourth, recombination is modelled in terms of
its outcomes in terms of inheritance intervals among parental genomes;
thus, we will have a child node (the recombinant; e.g. node $a$
Figure~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
paternal genomes which this child has inherited segments from.
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the evolutionary processes of genetic inheritance and
recombination, and is agnostic to the details of those processes.

@@

ARGs are most often thought of as a \emph{retrospective} structure however,
defined with respect to some set of (usually, contemporary) sample genomes.
In this case we are not explicitly interested in the
entire interval of genome descendant $u$ inherited from ancestor $v$,
but rather the subset of that interval that is \emph{ancestral to our
sample}.
Fig~\ref{fig-ancestry-resolution}B shows the graph that we obtain when
reduce the annotation intervals to those that are ancestral to the
sample (note that node $q$ is now ommitted because the MRCA of the sample
has already been found for the full genome at node $p$).
Tracking such
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
is much more useful~\citep{mcgill2013graphml,arenas2013importance}
and, indeed, all that is possible to
have information about when inferring ARGs from contemporary samples.
We refer to a gARG with inheritance annotations of this type as
``sample resolved''.
% Keeping track of resolved ancestral material is one of the distinguishing
% features of ARGs vs phylogenetic
% networks~\citep{mcgill2013graphml,arenas2013importance}, but
% is actually incompatible with the Griffiths gARG encoding (it does this
% implicitly, I guess)

@@

\section*{Computational efficiency and data interchange}
ARG inference methods have been successfully applied to vast datasets,
including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
from UK Biobank~\citep{bycroft2018genome}
and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
Even larger datasets are
available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
and it seems inevitable that ARGs will be inferred for them.
The efficiency with which we can store and analyse these inferred
ARGs is therefore of great importantance, and a key factor
in the ultimate success of downstream ARG-based methods.
Another key factor in how important ARG-based methods become
across the wide range of predicted application
% TODO more?
areas~\citep{rasmussen2014genome,harris2019database,
hejase2020summary,schaefer2021ancestral,
fan2022genealogical,wohns2022unified,harris2023using}
is how well inference methods and downstream analysis applications
interoperate. Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems of converting data from one programs
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
While current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package
% FIXME-  more explicit?
(e.g.\ Relate, ARGneedle)
this approach scales poorly in terms of software development
effort, and is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse
software ecosystem.
% FIXME not sure where this goes
Earlier efforts to standardise ARG interchange
did not succeed~\citep{cardona2008extended,mcgill2013graphml}.

@@

% FIXME this needs more work and doesn't really fit into the
% narrative right now, but we do need to make this point.
The key insight that makes the succinct tree sequence encoding
an efficient substrate for defining analysis algorithms is that
it allows us to generate the local trees along the genome
in a way that allows us to reason about the \emph{differences}
between those trees.
Sequentially generating the marginal trees along the genome
is also fundamental, and is necessary whenever we need to
perform calculations that are contingent on more than just the
isolated properties of an edge. \cite{kelleher2016efficient}
showed how all trees can be sequentially generated in
constant time per tree transition in a fully simplified gARG.
Furthermore, we can easily reason about how tree topologies
change (and stay the same), leading to efficient algorithms
for computing population genetic
statistics~\citep{kelleher2016efficient,kelleher2018efficient},
implementing the Li and Stephens
model~\citep{kelleher2019inferring,wohns2022unified}
and likely many more.

@@

\item It is unfortunate that we are adding yet another meaning to the
term ARG here, but all we're really doing is making the implicit
assumptions that others have made (i.e., describing the output of
Relate, tsinfer etc as ARGs) and making them concrete. We need to have a
term to describe all of these things, and since people are already using
ARG for this, we might as well make it rigorous. We can then move away
from the unfruitful discussion about which methods infer a ``true'' or
``real'' ARG (none do, really, under the original definitions) and
instead start discussing the \emph{properties} of these ARGs. (See
next point also.)

\item Terminology based around an inferred ancestry being a ``true''
ARG is impoverished and misleading. The only reasonable interpretation
of a ARG being true or not is whether it's a sample from the coalescent.
Most inference methods (and all that scale well) do not. In any case,
the coalescent is just one model, and a highly idealised one. It is
exceedingly unlikely that the actual ARG describing the ancestry
of any biological population is ``true'' ARG in this sense. The terminology is
impoverished because it reduces the possibilities available
to something either being an ARG or not, whereas there are many different
interesting properties of these structures that we should be able to
discuss and compare.

@@

@@
